---
title: "Trajectory Inference and VISION"
author: "Matthew Jones"
date: "`r BiocStyle::doc_date()`"
package: "`r BiocStyle::pkg_ver('BiocStyle')`"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Introduction to VISION}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

``` {r options, include=F, cache=F, results='hide', message=F}

knitr::opts_chunk$set(fig.align="center", cache=FALSE,error=FALSE,
                      fig.width=6,fig.height=6,autodep=TRUE,
                      out.width="600px", out.height="600px",
                      results="markup", echo=TRUE, eval=TRUE)

options(getClass.msg=FALSE)

set.seed(6473) ## for reproducibility

```

# Introduction

VISION can be used as a downstream tool for analyzing the biological significance of
trajectories. The tool interfaces with the objects output from [Dynverse](https://github.com/dynverse).
Briefly, Dynverse objects are general wrappers for any trajectory method, and VISION specifically
depends on two components in particular:

- The milestone network (essentially the topology of the trajectory produced from your
favorite trajectory method)
- The progression of cells along each edge in the milestone network. We expect that
each cell is only present on a single edge.

This amounts to a latent space, which describes relationships between cells, and can be used
for calculating signature consistency scores in the default pipeline.

## Installation

You should have a working installation of [VISION](https://github.com/YosefLab/VISION), [dyno](https://github.com/dynverse/dyno), and [tidyverse](https://github.com/tidyverse/tidyverse). 

```{r, collapse=F, message=T, eval=F}
devtools::install_github("YosefLab/VISION")
devtools::install_github("dynverse/dyno")
install.packages('tidyverse')
```

## Data

In this vignette, we'll be analyzing a set of ~5,000 cells during haematopoiesis ([Tusi et al, Nature 2018](https://www.nature.com/articles/nature25741)). 

# Workflow

We begin by loading in the data and perfomring perliminary scaling:

```{r, collapse=F, message=F, warning=F}
library(VISION)
library(dynmethods)
library(monocle)
library(DDRTree)
```

```{r, collapse=F, message=T, eval=F}

expr = as.matrix(read.table("data/hemato_counts.csv.gz", sep=',', header=T, row.names=1))
```

We'll now follow the [dyno vignette](https://github.com/dynverse/dyno) in the github README. You'll need to make sure you have [Docker](https://docs.docker.com/install/) installed. Although we use Monocle2 here, there is a wide variety of methods that can be used, as listed on the [Dynmethods page](https://github.com/dynverse/dynmethods). Note that for Monocle, we are feeding in the raw counts, not scaled as you would when passing data into VISION.

```{r, collapse=F, message=T, eval=F}

model <- dynmethods:::run_monocle(t(expr), reduction_method="DDRTree", norm_method="log")
```

Now, with this model saved we can pass it to VISION via the `latentTrajectory` parameter. The model will automatically be analzyed with the same signatures passed to the FastProjectR object, and with this model, the `Trajectory` tab will be activated in the FastProjectR output report. 

``` {r, collapse=F, eval=F}
covar = read.table("data/hemato_covariates.txt", sep='\t', header=T, row.names=1)
covar = covar[rownames(expr),]
vis <- VISION(expr, c("data/h.all.v5.2.symbols.gmt"), projection_genes = rownames(expr), 
            meta=covar, latentTrajectory = model, sig_norm_method="znorm_columns")
```

# Session Info
```{r, collapse=F, message=T}
sessionInfo()
```