---
title: "Incorporating a Trajectory Model"
package: "`r BiocStyle::pkg_ver('BiocStyle')`"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Incorporating a Trajectory Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

``` {r options, include=F, cache=F, results='hide', message=F}

knitr::opts_chunk$set(fig.align="center", cache=FALSE,error=FALSE,
                      fig.width=6,fig.height=6,autodep=TRUE,
                      out.width="600px", out.height="600px",
                      results="markup", echo=TRUE, eval=TRUE)

options(getClass.msg=FALSE)

set.seed(6473) ## for reproducibility

```

# Introduction

*Note: this feature is deprecated in VISION 3.0+.*

To compute Signature consistency scores, VISION needs a cell-cell similarity metric in order to
build the neighborhood graph.  By default, VISION runs PCA and uses euclidean distances in the reduced
space for this purpose.  However, if the cells' variation is better modeled by a trajectory (either
linear or tree-like), then distances along the trajectory structure can be used instead.

Many tools exist to perform trajectory inference in single cells.  The [Dynverse](https://github.com/dynverse)
project has developed wrappers for 59 recently-published methods and VISION is able to accept
the wrapped output of these methods as an input trajectory model.

Within the Dynverse trajectory object, VISION depends on two components:

- The **milestone network** (essentially the topology of the trajectory produced from your
favorite trajectory method)
- The **progression** of cells along each edge in the milestone network. We expect that
each cell is only present on a single edge.

## Installation

You should have a working installation of [VISION](https://github.com/YosefLab/VISION), [dyno](https://github.com/dynverse/dyno), and [tidyverse](https://github.com/tidyverse/tidyverse).

```{r, collapse=F, message=T, eval=F}
devtools::install_github("YosefLab/VISION")
devtools::install_github("dynverse/dyno")
install.packages('tidyverse')
```

## Data

In this vignette, we'll be analyzing a set of ~5,000 cells during haematopoiesis ([Tusi et al, Nature 2018](https://www.nature.com/articles/nature25741)).

# Workflow

We begin by loading in the data and requisite libraries:

```{r, collapse=F, message=F, warning=F, eval=F}
library(VISION)
library(dyno)
library(tidyverse)
```

```{r, collapse=F, message=T, eval=F}
counts = as.matrix(read.table("data/hemato_counts.csv.gz", sep=',', header=T, row.names=1))
```

We'll now follow the [dyno vignette](https://github.com/dynverse/dyno) in the github README. You'll need to make sure you have [Docker](https://docs.docker.com/install/) installed. Although we use Slingshot here, there is a wide variety of methods that can be used, as listed on the [Dynmethods page](https://github.com/dynverse/dynmethods).

Before running the model, we'll need to perform some gene filtering. We first only select the genes used in the original Tusi et al study, then for further efficiency, we further filter using VISION's Fano filtering.

```{r, collapse=F, message=T, eval=F}
k.genes = read.table("data/bBM.filtered_gene_list.paper.txt.gz", sep=',')[,1]
filt.counts = counts[k.genes, ]

f.genes = VISION:::filterGenesFano(filt.counts)
filt.counts = filt.counts[f.genes, ]

```

Now, resuming the workflow recommended on the dyno vignette,

```{r, collapse=F, message=T, eval=F}
umis <- colSums(counts)
scale.factor <- median(umis)
expr <- t(t(counts) / umis) * scale.factor
filt.expr <-  log(1 + expr[f.genes,])

dataset <- wrap_expression(
  counts = t(filt.counts),
  expression = t(filt.expr)
)

model <- infer_trajectory(dataset, "projected_slingshot", verbose=T)

model <- model %>% add_dimred(dyndimred::dimred_mds, expression_source = dataset$expression)

plot_dimred(
  model,
  expression_source = dataset$expression
)
```

Now, with this model saved we can pass it to VISION via the `latentTrajectory` parameter. The model will automatically be analyzed with the same signatures passed to the VISION object, and with this model, the `Trajectory` tab will be activated in the VISION output report.

```{r, collapse=F, eval=F}
meta <- read.table("data/hemato_covariates.txt.gz", sep='\t', header=T, row.names=1)
meta <- meta[colnames(expr), -1]

vis <- Vision(expr, signatures = c("data/h.all.v5.2.symbols.gmt"),
            meta = meta, latentTrajectory = model)

vis <- addProjection(vis, "MDS", model$dimred[,c(1,2)])

vis <- analyze(vis)
```


# Session Info
```{r, collapse=F, message=T}
sessionInfo()
```
