---
title: "Trajectory Inference and FastProjectR"
author: "Matthew Jones"
date: "`r BiocStyle::doc_date()`"
package: "`r BiocStyle::pkg_ver('BiocStyle')`"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Introduction to FastProjectR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

``` {r options, include=F, cache=F, results='hide', message=F}

knitr::opts_chunk$set(fig.align="center", cache=FALSE,error=FALSE,
                      fig.width=6,fig.height=6,autodep=TRUE,
                      out.width="600px", out.height="600px",
                      results="markup", echo=TRUE, eval=TRUE)

options(getClass.msg=FALSE)

set.seed(6473) ## for reproducibility

```

# Introduction

FastProjectR can be used as a downstream tool for analyzing the biological significance of
trajectories. The tool interfaces with the objects output from [Dynverse](https://github.com/dynverse).
Briefly, Dynverse objects are general wrappers for any trajectory method, and Fastproject specifically
depends on two components in particular:

- The milestone network (essentially the topology of the trajectory produced from your
favorite trajectory method)
- The progression of cells along each edge in the milestone network. We expect that
each cell is only present on a single edge.

This amounts to a latent space, which describes relationships between cells, and can be used
for calculating signature consistency scores in the default pipeline.

## Installation

You should have a working installation of [FastProjectR](https://github.com/YosefLab/FastProjectR), [dyno](https://github.com/dynverse/dyno), and [tidyverse](https://github.com/tidyverse/tidyverse). 

```{r, collapse=F, message=T, eval=F}
devtools::install_github("YosefLab/FastProjectR")
devtools::install_github("dynverse/dyno")
install.packages('tidyverse')
```

## Data

In this vignette, we'll be analyzing a set of ~5,000 cells during haematopoiesis ([Tusi et al, Nature 2018](https://www.nature.com/articles/nature25741)). 

# Workflow

We begin by loading in the data and perfomring perliminary scaling:

```{r, collapse=F, message=F, warning=F}
library(FastProjectR)
library(dyno)
library(tidyverse)
```

```{r, collapse=F, message=T, eval=F}

expr = as.matrix(read.table("data/hemato_counts.csv.gz", sep=',', header=T, row.names=1))

# perform preliminary scaling
n.umi = median(colSums(expr))
expr = apply(expr, 2, function(x) (x * n.umi) / (sum(x) + 1))

# get informative genes
f.genes = filterGenesFano(expr)
expr = expr[f.genes,]

# log-normalize expression matrix
n.expr = log2(expr)
```

We'll now follow the [dyno vignette](https://github.com/dynverse/dyno) in the github README. You'll need to make sure you have [Docker](https://docs.docker.com/install/) installed. Although we use Slingshot here, there is a wide variety of methods that can be used, as listed on the [Dynmethods page](https://github.com/dynverse/dynmethods).

```{r, collapse=F, message=T, eval=F}
task <- wrap_expression(
  counts = t(expr),
  expression = t(n.expr)
)

start_dynmethods_docker()

model %<-% infer_trajectory(task, ti_slingshot())

saveRDS(model, file="hemato_traj_model.RDS")
```

Now, with this model saved we can pass it to FastProjectR via the `latentTrajectory` parameter. The model will automatically be analzyed with the same signatures passed to the FastProjectR object, and with this model, the `Trajectory` tab will be activated in the FastProjectR output report. 

# Session Info
```{r, collapse=F, message=T}
sessionInfo()
```