---
title: "Trajectory Inference and VISION"
author: "Matthew Jones"
date: "`r BiocStyle::doc_date()`"
package: "`r BiocStyle::pkg_ver('BiocStyle')`"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Introduction to VISION}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

``` {r options, include=F, cache=F, results='hide', message=F}

knitr::opts_chunk$set(fig.align="center", cache=FALSE,error=FALSE,
                      fig.width=6,fig.height=6,autodep=TRUE,
                      out.width="600px", out.height="600px",
                      results="markup", echo=TRUE, eval=TRUE)

options(getClass.msg=FALSE)

set.seed(6473) ## for reproducibility

```

# Introduction

VISION can be used as a downstream tool for analyzing the biological significance of
trajectories. The tool interfaces with the objects output from [Dynverse](https://github.com/dynverse).
Briefly, Dynverse objects are general wrappers for any trajectory method, and VISION specifically
depends on two components in particular:

- The milestone network (essentially the topology of the trajectory produced from your
favorite trajectory method)
- The progression of cells along each edge in the milestone network. We expect that
each cell is only present on a single edge.

This amounts to a latent space, which describes relationships between cells, and can be used
for calculating signature consistency scores in the default pipeline.

## Installation

You should have a working installation of [VISION](https://github.com/YosefLab/VISION), [dyno](https://github.com/dynverse/dyno), and [tidyverse](https://github.com/tidyverse/tidyverse). 

```{r, collapse=F, message=T, eval=F}
devtools::install_github("YosefLab/VISION")
devtools::install_github("dynverse/dyno")
install.packages('tidyverse')
```

## Data

In this vignette, we'll be analyzing a set of ~5,000 cells during haematopoiesis ([Tusi et al, Nature 2018](https://www.nature.com/articles/nature25741)). 

# Workflow

We begin by loading in the data and requisite libraries:

```{r, collapse=F, message=F, warning=F, eval=F}
library(VISION)
library(dyno)
library(tidyverse)
```

```{r, collapse=F, message=T, eval=F}

counts = as.matrix(read.table("data/hemato_counts.csv.gz", sep=',', header=T, row.names=1))

```

We'll now follow the [dyno vignette](https://github.com/dynverse/dyno) in the github README. You'll need to make sure you have [Docker](https://docs.docker.com/install/) installed. Although we use Monocle2 here, there is a wide variety of methods that can be used, as listed on the [Dynmethods page](https://github.com/dynverse/dynmethods). 

Before running the model, we'll need to perform some gene filtering. We first only select the genes used in the original Tusi et al study, then for further efficiency, we further filter using VISION's Fano filtering.

```{r, collapse=F, message=T, eval=F}
k.genes = read.table("data/bBM.filtered_gene_list.paper.txt.gz", sep=',')[,1]
filt.counts = counts[k.genes,]

f.genes = VISION:::filterGenesFano(filt.counts)
filt.counts = filt.counts[f.genes,]

```

Now, resuming the workflow recommended on the dyno vignette,

```{r, collapse=F, message=T, eval=F}

scale.factor = median(colSums(counts))
expr = apply(counts, 2, function(x) (x * scale.factor) / (sum(x) + 1))
filt.expr = log(1 + expr[f.genes,])

dataset = wrap_expression(
  counts = t(filt.counts),
  expression = t(filt.expr)
)

model <- infer_trajectory(dataset, "projected_slingshot", verbose=T)

model <- model %>% add_dimred(dyndimred::dimred_mds, expression_source = dataset$expression)
plot_dimred(
  model, 
  expression_source = dataset$expression
)

```

Now, with this model saved we can pass it to VISION via the `latentTrajectory` parameter. The model will automatically be analzyed with the same signatures passed to the FastProjectR object, and with this model, the `Trajectory` tab will be activated in the VISION output report. 

```{r, collapse=F, eval=F}
covar = read.table("data/hemato_covariates.txt.gz", sep='\t', header=T, row.names=1)
covar = covar[colnames(expr), -1]
vis <- Vision(expr, 
            c("data/h.all.v5.2.symbols.gmt"), 
            projection_genes = f.genes, 
            meta=covar, 
            latentTrajectory = model, 
            sig_norm_method="znorm_columns")
vis <- addProjection(vis, "MDS", model$dimred[,c(1,2)])

vis <- analyze(vis)
```



# Session Info
```{r, collapse=F, message=T}
sessionInfo()
```
